### 2021-04-26 conclusion.tex
Update conclusion.tex
Final version.

While our work focuses on \emph{binary} session types, it is worth to discuss related work on Multiparty Session Types (MPST). The line of work on MPST starts with Honda~\etal~\cite{hondayoshida08}, which guarantees deadlock freedom \emph{within a single session}, but not for session interleaving.
Bettini~\etal~\cite{bettinicoppo08} follow a technique similar to Kobayashi's for MPST.
The main difference with our work is that we associate priorities with communication actions, where Bettini~\etal~\cite{bettinicoppo08} associate them with channels.
Carbone and Montesi~\cite{carbonemontesi13} combine MPST with choreographies and obtain a formalism that satisfies deadlock freedom by design.
Deni\'{e}lou and Yoshida~\cite{DenielouY13} introduce \emph{multiparty compatibility} which generalises duality in binary session types. They synthesise safe and deadlock-free global types from local types leveraging LTSs and communicating automata.
Castellani~\etal~\cite{CastellaniDGH20} guarantee lock freedom, a stronger property than deadlock freedom, for MPST with \emph{internal delegation}, where participants in the same session are allowed to delegate tasks to each other, and internal delegation is captured by the global type.
Scalas and Yoshida~\cite{scalasyoshida19} provide a revision of the foundations for MPST, and offer a less complicated and more general theory, by removing duality/consistency. The type systems is parametric and type checking is decidable, but allows for a novel integration of model checking techniques. More protocols and processes can be typed and are guaranteed to be free of deadlocks.

Gay~\etal~\cite{gaynagarajan03} and Vasconcelos~\etal~\cite{vasconcelosravara04,vasconcelosgay06} introduce the first functional language with session types. Such works did not guarantee deadlock freedom until GV~\cite{lindleymorris15,wadler14}.
Toninho~\etal~\cite{toninhocaires12} present a translation of simply-typed $\lambda$-calculus into session-typed $\pi$-calculus, but their focus is not on deadlock freedom.

@@

\subsubsection*{Future work}
Our formalism so far only captures the core of GV. In future work, we plan to explore recursion, following Lindley and Morris~\cite{lindleymorris16} and Padovani and Novara~\cite{padovaninovara15}, and sharing, following Balzer and Pfenning~\cite{balzerpfenning17} or Voinea~\etal~\cite{VoineaDG19}.
% or in the lines of Kobayashi and Laneve for unbounded networks~\cite{KL17}, which is a refinement of previous work by Kobayashi.

>>>>

While our work focuses on \emph{binary} session types, it is worth to discuss related work on Multiparty Session Types (MPST). The line of work on MPST starts with Honda~\etal~\cite{hondayoshida08}, which guarantees deadlock freedom \emph{within a single session}, but not for session interleaving.
Bettini~\etal~\cite{bettinicoppo08} follow a technique similar to Kobayashi's for MPST.
The main difference with our work is that we associate priorities with communication actions, where Bettini~\etal~\cite{bettinicoppo08} associate them with channels.
Carbone and Montesi~\cite{carbonemontesi13} combine MPST with choreographies and obtain a formalism that satisfies deadlock freedom.
Deni\'{e}lou and Yoshida~\cite{DenielouY13} introduce \emph{multiparty compatibility} which generalises duality in binary session types. They synthesise safe and deadlock-free global types from local types leveraging LTSs and communicating automata.
Castellani~\etal~\cite{CastellaniDGH20} guarantee lock freedom, a stronger property than deadlock freedom, for MPST with \emph{internal delegation}, where participants in the same session are allowed to delegate tasks to each other, and internal delegation is captured by the global type.
Scalas and Yoshida~\cite{scalasyoshida19} provide a revision of the foundations for MPST, and offer a less complicated and more general theory, by removing duality/consistency. The type systems is parametric and type checking is decidable, but allows for a novel integration of model checking techniques. More protocols and processes can be typed and are guaranteed to be free of deadlocks.

Neubauer and Thiemann~\cite{neubauert04} and Vasconcelos~\etal~\cite{vasconcelosravara04,vasconcelosgay06} introduce the first functional language with session types. Such works did not guarantee deadlock freedom until GV~\cite{lindleymorris15,wadler14}.
Toninho~\etal~\cite{toninhocaires12} present a translation of simply-typed $\lambda$-calculus into session-typed $\pi$-calculus, but their focus is not on deadlock freedom.

@@

\subsubsection*{Future work}
Our formalism so far only captures the core of GV. In future work, we plan to explore recursion, following Lindley and Morris~\cite{lindleymorris16} and Padovani and Novara~\cite{padovaninovara15}, and sharing, following Balzer and Pfenning~\cite{balzerpfenning17} or Voinea~\etal~\cite{VoineaDG19}.

### 2021-04-26 introduction.tex
Final version.

\begin{document}
\section{Introduction}
Session types~\cite{honda93,takeuchihonda94,hondavasconcelos98} are types for protocols. Regular types ensure functions are used according to their specification. Session types ensure \emph{communication channels} are used according to their protocols. Session types have been studied in many settings. For instance, in the $\pi$-calculus~\cite{honda93,takeuchihonda94,hondavasconcelos98}, a foundational calculus for communication and concurrency, and in concurrent $\lambda$-calculi~\cite{gayvasconcelos12}, including the focus of our paper: Good Variation~\cite[GV]{wadler15,lindleymorris15}.

>>>>

\begin{document}
\section{Introduction}
Session types~\cite{honda93,takeuchihonda94,hondavasconcelos98} are types for protocols. Regular types ensure functions are used according to their specification. Session types ensure \emph{communication channels} are used according to their protocols. Session types have been studied in many settings. For instance, in the $\pi$-calculus~\cite{honda93,takeuchihonda94,hondavasconcelos98}, a foundational calculus for communication and concurrency, and in concurrent $\lambda$-calculi~\cite{gayvasconcelos10}, including the focus of our paper: Good Variation~\cite[GV]{wadler15,lindleymorris15}.

### 2021-04-26 priority-gv.tex
Final version.

We present Priority GV (PGV), a session-typed functional language based on GV~\cite{wadler15,lindleymorris15} which uses priorities \`{a} la Kobayashi and Padovani~\cite{kobayashi06,padovaninovara15} to enforce deadlock freedom.
Priority GV offers a more fine-grained analysis of communication structures, and by separating channel creation form thread spawning it allows cyclic structures.
We illustrate this with two programs in PGV, \cref{ex:cycle,ex:deadlock}. Each program contains two processes---the main process, and the child process created by $\tm{\spawn}$---which communicate using \emph{two} channels. The child process receives a unit over the channel $\tm{x}/\tm{x'}$, and then sends a unit over the channel $\tm{y}/\tm{y'}$. The main process does one of two things:
\begin{enumerate*}[label= (\alph*) ]
\item in \cref{ex:cycle}, it sends a unit over the channel $\tm{x}/\tm{x'}$, and then waits to receive a unit over the channel $\tm{y}/\tm{y'}$;
\item in \cref{ex:deadlock}, it does these in the opposite order, which results in a deadlock.
\end{enumerate*}
PGV is more expressive than GV: \cref{ex:cycle} is typeable and guaranteed deadlock-free in PGV, but is not typeable in GV~\cite{wadler14} and not guaranteed deadlock-free in GV's predecessor~\cite{gayvasconcelos12}. We believe PGV is a non-conservative \emph{extension} of GV, as CP can be embedded in a Kobayashi-style system~\cite{dardhaperez15extended}.

>>>>

We present Priority GV (PGV), a session-typed functional language based on GV~\cite{wadler15,lindleymorris15} which uses priorities \`{a} la Kobayashi and Padovani~\cite{kobayashi06,padovaninovara15} to enforce deadlock freedom.
Priority GV offers a more fine-grained analysis of communication structures, and by separating channel creation form thread spawning it allows cyclic structures.
We illustrate this with two programs in PGV, \cref{ex:cycle,ex:deadlock}. Each program contains two processes---the main process, and the child process created by $\tm{\spawn}$---which communicate using \emph{two} channels. The child process receives a unit over the channel $\tm{x}/\tm{x'}$, and then sends a unit over the channel $\tm{y}/\tm{y'}$. The main process does one of two things:
\begin{enumerate*}[label= (\alph*) ]
\item in \cref{ex:cycle}, it sends a unit over the channel $\tm{x}/\tm{x'}$, and then waits to receive a unit over the channel $\tm{y}/\tm{y'}$;
\item in \cref{ex:deadlock}, it does these in the opposite order, which results in a deadlock.
\end{enumerate*}
PGV is more expressive than GV: \cref{ex:cycle} is typeable and guaranteed deadlock-free in PGV, but is not typeable in GV~\cite{wadler14} and not guaranteed deadlock-free in GV's predecessor~\cite{gayvasconcelos10}. We believe PGV is a non-conservative \emph{extension} of GV, as CP can be embedded in a Kobayashi-style system~\cite{dardhaperez15extended}.
